---
title: "Homework Information"
author: "MATH 456: Spring 2016"
output: html_document
---

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](index.html) [[Schedule]](Schedule.html) [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)  [[Google Drive]](https://drive.google.com/a/mail.csuchico.edu/folderview?id=0B83Z8_sNw3KPcnVrYzVFRHUtcHM&usp=sharing)



```{r, include = FALSE}
library(readr)
#rubric <- read_csv(file.path("admin", "hw_rubric.csv"))
rubric <- read_csv("admin/hw_rubric.csv")
```
## Posted Solutions
This is where I will post homework solutions (full and partial), and some helper code.

* Data Wrangling Solutions [Ch3_solutions](notes/solutions/ch3_solutions.html)
* Simple Linear regression [SLR_solutions](notes/solutions/SLR_solutions.html)
* Multiple Linear regression [MLR_solutions](notes/solutions/MLR_solutions.html)
* Plotting help for spline predictions [Lec04_solutions](notes/solutions/Lec04_solutions.html)
* My model building [VariableSelection_solutions](notes/solutions/VariableSelection_solutions.html)



## How to submit homework
* Homework will be submitted through BBLearn approximately every other week.
* The file name of your submission should follow this format: _userid_assigment_
    - For example my markdown document for chapter 3 would be _rdonatello_ch3.Rmd_
* Double check the notes and blackboard learn to see what file exactly is to be uploaded. 
    - rmd/word/html/pdf 
* The requested file should be submitted via BBLearn by the due date. 
  Late homework is generall not accepted. 
* Grading will be video recorded. The markedup document and video recording 
  will be put into your personal MATH456 Google Drive folder that you shared with me. 

      
## How homework will be graded
* If you submit a code file then I will download your code file, and compile
  it on my machine. What is generated is what I grade. 
  Make sure it looks good before you upload it. 
* A random set of 1-3 problems will be chosen to be graded. 
* Grading will be done by using the rubric below. 
* The full assignment is worth 10 pts.
    - 3 criteria worth up to 3 pts. 
    - 1 pt for a complete assignment. 

## Grading Rubric
```{r results = 'asis', echo = FALSE}
knitr::kable(rubric)
```