---
title: 'Lec 08: Logistic Regression'
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
  pdf_document: default
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
library(xtable); library(dplyr)
options(xtable.comment = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center') 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 11 Overview]](../wk11.html) [[HW Info]](../HW_Info.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)


# Introduction 

## Assigned Reading & additional references




## Going further
When your outcome has more than one level and you want to build a regression
model to assess the impact a specific variable (or set of variables) has on 
the levels of this outcome variable, you would need to turn to more
generalized linear models such as: 
* Multinomial distribution for a nominal outcome
    - http://www.ats.ucla.edu/stat/r/dae/mlogit.htm
* Ordinal logistic regression


[[top]](lec08_LogisticRegression.html)

# On Your Own

##### On Your Own

    
    
    