---
title: "Lec 02: Linear Regression Analysis"
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
  pdf_document: default
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);
opts_chunk$set(warning=FALSE, message=FALSE) 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

# Assigned Reading
Afifi: Chapters 6-7

# Simple Regression and Correlation (_Afifi Ch 6_)

## Aims
* Describe the relationship between an independent variable X and a continuous
  dependent variable $Y$ as a straight line. The textbook discusses two cases: 
    - Fixed-$X$: values of $X$ are preselected by investigator
    - Variable-$X$: have random sample of $(X,Y)$ values
    - Calculations are the same, 
* Draw inferences regarding this relationship
* Predict value of $Y$ for a given value of $X$

## Mathmatical Model
* The mean of $Y$ values at any given $X$ is $\alpha + \beta X$
* The variance of $Y$ values at any $X$ is $\sigma^2$ (same for all X)
* $Y$ values are normally distributed at any given $X$ (need for inference)



##### On Your Own: 
Write your responses in a new Markdown file named *userid_ch6.rmd*.



# Multiple Regression and Correlation (_Afifi Ch 7_)


##### On Your Own
Write your responses in a new Markdown file named *userid_ch7.rmd*.


