---
title: 'Lec 07: Categorical Data Analysis'
author: "MATH 456 - Spring 2016"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
library(xtable); library(dplyr)
options(xtable.comment = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center') 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 10 Overview]](../wk10.html) [[HW Info]](../HW_Info.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)

# Assigned Reading & additional references
* OpenIntro Lab on categorical data [[HTML]](notes/OI_Inf_for_categorical_data/html)
* OpenIntro Statistics [Free PDF Textbook](https://www.openintro.org/stat/textbook.php) Chapter 6 (6.1-6.4)
* Afifi Chapter 12 (Logistic Regression)

**Additional references**

* Data vis tutorial http://norcalbiostat.github.io/R-Bootcamp/labs/Data_Visualization_Tutorial_Full.html
* Cookbook for R http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/

# Introduction 
Up until now we have been analyzing continuous outcomes. We will now turn our
focus to methods to analyze categorical data. This will allow us to answer 
questions like the following:

* What proportion of the American public approves of the job the Supreme Court
  is doing?
* The Pew Research Center conducted a poll about support for the 2010 health 
  care law, and they used two forms of the survey question. Each respondent
  was randomly given one of the two questions. What is the difference in the
  support for respondents under the two question orderings?

The methods you learned in previous classes will be useful in these settings. 
For example, sample proportions are well characterized by a nearly normal 
distribution when certain conditions are satisfied, making it possible 
to employ the usual CI and HT tools. In other instances, such as those with 
contingency tables or when sample size conditions are not met, we will use a 
different distribution, though the core ideas remain the same.

Be sure to referece the categorical data sections of the [Data Visualization Tutorial](http://norcalbiostat.github.io/R-Bootcamp/labs/Data_Visualization_Tutorial_Full.html) or 
the [Cookbook for R](http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/) for
more information and guidance on how to plot categorical data correctly. 

#### Spam Data
This set of lecture notes uses the `spam` data set. 
```{r}
email <- read.delim("C:/Dropbox/CSUC/Data/email.txt", header=TRUE, sep="\t")
```

Two categorical variables of current interest are

* `spam` (0/1 binary indicator if a an email is flagged as spam). Converted into a Ham/Spam factor variable. 
* `number` categorical variable describing the size of the numbers contained in the email.  
    - `none`: No numbers 
    - `small`: Only values under 1 million
    - `big`: A value of 1 million or more

# Review: Inference on a single proportion 
This section is considered review and will not be covered directly in 
class. If you have never analyzed proportions in R or it has been a while 
since you took MATH 315 it is advised that you do not skip this section. 

## Inference for a single proportion
* Open Intro: Chapter 6.1
* Complete the OpenIntro Lab on categorical data [[HTML]](notes/OI_Inf_for_categorical_data/html)

**You must have the `openintro` package installed & loaded to have access 
to the custom `inference()` function used in the lab.**

### Example: Examining the frequency of number sizes in emails

#### One-way Tables
A table for a single variable is called a _frequency table_. The values 
displayed represent the number of records in the data set that fall into that
particular category.
```{r}
table(email$number)
```

If we replaced the counts with percentages or proportions, the table would be
called a _relative frequency table_.

```{r}
prop.table(table(email$number))
```

We make this output more human readable as percentages by rounding the results
and multplying by 100.
```{r}
round(prop.table(table(email$number))*100,2)
```

Out of the 3921 emails in this data set, 545 (13.9%) contain big numbers, 
2827 (72.1%) contain small numbers, and 549 (14.0%) contain no numbers. 

#### One variable barcharts

The most common method to display frequencies in a graphical manner is
with a _bar chart_ (aka barplot or bar graph). One bar per distinct
category with the height of the bar representing the frequency (or percent)
of the data that fall into that category. 

```{r, fig.align='center', fig.width=4, fig.height=4}
ggplot(email, aes(number)) + geom_bar() + ylab("Frequency") + xlab("Size of number in email") + 
  ggtitle("Frequency of emails by size of the number") 
```

#### Research Question: Do more than half of the emails flagged as spam contain numbers?
We are only interested in the emails that are flagged spam, and that 
contain any number, so let's do a little data cleaning first. 

```{r}
spam.num <- email %>% 
              filter(spam==1) %>%  
              mutate(hasnum = ifelse(number %in% c("big", "small"), 1, 0))
# Confirm recode
table(spam.num$number, spam.num$hasnum, useNA="always")
```

* What is the proportion of spam emails that contain numbers? 
Since we created `hasnum` as a binary indicator, we can calculate the
sample proportion as the sample mean. 
```{r}
mean(spam.num$hasnum)
```
* Construct a 99% CI for the proportion of spam emails with numbers. 
Since this proportion follows the normal model we can conduct a t.test. 
```{r}
t.test(spam.num$hasnum, alternative = "two.sided", conf.level=.99)
```

Over half (59.4%, 95%CI: 52.7, 66.0) of emails that are flagged spam
contain numbers (p<.0001). This is statistically significantly larger
than half. 


[[top]](lec07_CategoricalDataAnalysis.html)

# Inference for two categorical variables

## Two way (two variable) tables

```{r}
tab <- table(email$spam, email$number)
tab
```

#### Cell percents
A simple `prop.table()` shows the cell percents. 
All cells add up to 1. The `mosaicplot` is a visual
representation of these percentages. The larger the 
box the larger the proportion. 
```{r}
prop.table(tab)
mosaicplot(tab)
```

We will come back to these mosaicplots later. 


#### Row percents
Here, the percents add up to 1 across the rows. 
The reference group (denominator) is the row category.  
```{r}
round(prop.table(tab, 1)*100, 2)
```
* 13.9% _of non-spam emails_ contain big numbers
* 40.6% _of spam emails_ contain no numbers
* 74.8% _of non-spam emails_ contain small numbers. 

#### Base graphics
Does "ok", but it depends on how you set up your table (rows vs columns),
and you  have to do more work to get a reasonable legend and y axis. 
```{r}
barplot(tab, beside=TRUE, legend=rownames(tab))
```

#### GGplot2
Requires the data to be aggregated first, then specify the heights
of the bars as a new variable. 
```{r}
library(reshape2)
melted_tab <- melt(tab)
colnames(melted_tab) <- c("Spam", "Number", "count")
melted_tab
ggplot(melted_tab, aes(x = factor(Spam), y= count, fill = Number)) + 
  geom_bar(stat="identity", width=.5, position = "dodge")  + 
  scale_x_discrete("Type of Email", labels=c("Ham", "Spam")) + 
  ylab("Frequency") 
```


#### Column percents
Here, the percents add up to 1 down the columns. 
The reference group (denominator) is the column category.  
```{r}
round(prop.table(tab, 2)*100, 2)
```
* 90.8% _of emails with big numbers_ are not spam
* 27.1% _of emails with no numbers_ are spam
* 94.1% _of emails with small numbers_ are not spam

```{r}
ggplot(melted_tab, aes(x = Number, y= count, fill = factor(Spam))) + 
  geom_bar(stat="identity", width=.5, position = "dodge")  + 
  xlab("Size of number in email") + ylab("Frequency") + 
  scale_fill_discrete("Type of Email", labels=c("Ham", "Spam")) 
```

#### Summary / Take home message
It is very important to be clear as to what comparison you (or the researcher)
are interested in making. Sometimes both directions are equally important, 
but often there is one primary direction that is of interest. 


## Difference of two proportions
* OpenIntro Section 6.2

Now let's consider comparisons of proportions in two independent samples. 

**Example**: Comparison of proportions of head injuries sustained in auto 
accidents by passengers wearing seat belts to those not wearing seat belts. 
You may have already guessed the form of the estimate: $\hat{p}_1 - \hat{p}_2$.

We are not going to go in depth into the calculations for the test statistic 
for a test of the difference in proportions. The OpenIntro textbook explains 
the assumptions and equations very well. Instead we are going to see how to 
use R to perform these calculations for us. 

Since the sample proportion can be calculated as the mean of a binary 
indicator variable, we can use the same `t.test` function in R
to conduct a hypothesis test and create a confidence interval.

### Tests of proportions using case level data
Do emails that contain numbers more likely to be spam? 

```{r}
email <- email %>% mutate(hasnum =ifelse(number %in% c("big", "small"), 0, 1))
t.test(spam~hasnum, data=email)
```



### Tests of proportions in R using summary numbers only
Sometimes you only have a summary table of numbers to go on, not the raw data. 
We can use the `prop.test()` function in R to conduct a test of equal proportions. 

This function calculates a test statistic and a p-value using a $\chi^2$ distribution. 

* The $\chi^2$ distribution is mathematically related to many other distributions. 
    * Specific of note here is that a $Z^{2}$ distribution has a $\chi^{2}_{1}$ distribution. 
    * Other relationships that are very useful: 
        * https://onlinecourses.science.psu.edu/stat414/node/154
        * https://en.wikipedia.org/wiki/Chi-squared_distribution#Relation_to_other_distributions

#### Example 1: Spam vs Ham
Recall that 218 out of 367 (59%) of all emails that were flagged spam contained numbers. 
```{r}
table((email$number !="none"), email$spam, dnn=c("Contain numbers", "Spam"))
round(prop.table(table((email$number !="none"), email$spam, dnn=c("Contain numbers", "Spam")), 2)*100,2)
```
We can test if that proportion is significantly more than half by calling `prop.test()`. 

```{r}
prop.test(x=218, n=367, p=.5, alternative="greater")
```

#### Example 2: Are mammograms helpful?
Test whether there was a difference in breast cancer deaths in the mammogram 
and control groups. By entering in $x$ and $n$ as vectors we can test 
equivelance of these two proportions. The assumptions for using the normal model 
for this test have been discussed in detail in the textbook. 

```{r}
prop.test(x=c(500, 505), n=c(44925, 44910))
```







[[top]](lec07_CategoricalDataAnalysis.html)

# Chi-Squared tests for contingency tables. 

prop.test(table(email$number, email$spam))

mosaicplot(tab, shade=TRUE)

## Testing Equality of proportions

## Testing Goodness of Fit

## Testing Independence

## Comments on Fishers Exact test


[[top]](lec07_CategoricalDataAnalysis.html)

# Logistic Regression


## Going further
When your outcome has more than one level and you want to build a regression
model to assess the impact a specific variable (or set of variables) has on 
the levels of this outcome variable, you would need to turn to more
generalized linear models such as: 
* Multinomial distribution for a nominal outcome
    - http://www.ats.ucla.edu/stat/r/dae/mlogit.htm
* Ordinal logistic regression


[[top]](lec07_CategoricalDataAnalysis.html)

# On Your Own
For all hypothesis tests you must: 

- Clearly state what your null and alternative hypothesis.  
- Discuss your assumptions
- Show your R code to conduct the test and any data management required prior to the test.
- Write the conclusion of the statistical test in context of the problem. 

##### On Your Own
1. Using the `survey` data set contained in the `MASS` library, recode the exercise 
   variable into a dichotomous indicator of any exercise. Test the hypothesis that more 
   than 75% of the students in the sample exercise at some level.
2. Using the student `survey` data where the variable `smoker` is a binary indicator 
   of smoking, test if males smoke more than females. _Hint: check your output 
   carefully and consider if the difference was taken in the desired direction_
<!--
```{r}
library(MASS)
survey$yessmoker <- ifelse(survey$Smoke !="Never", 1, 0)
t.test(yessmoker ~ Sex , data=survey, alternative = "less")
```
--->
    
    
    
    