---
title: 'Lec 07: Discriminant Analysis'
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
  pdf_document: default
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
library(xtable); library(dplyr)
options(xtable.comment = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 9 Overview]](../wk09.html) [[HW Info]](../HW_Info.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)

# Assigned Reading & additional references

* Afifi Chapter 11

**Additional references**

* http://www.statmethods.net/advstats/discriminant.html
* http://rstudio-pubs-static.s3.amazonaws.com/35817_2552e05f1d4e4db8ba87b334101a43da.html
* https://www.youtube.com/watch?v=s8pvp2Ctxfc
* http://www.r-bloggers.com/computing-and-visualizing-lda-in-r/


# Introduction t
Discriminant analysis 


* Construct a rule to classify individuals into one of two or more groups.
* Compute probability of belonging to a given group.


[[top]](lec06_DimensionReduction.html)

# On Your Own
##### On Your Own

    
    
    
    