---
title: 'Lec 04: Indicator variables and Splines'
author: "MATH 456 - Spring 2016"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 5 Overview]](../wk05.html) [[HW Rubric]](../admin/rubric.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)  [[Google Drive]](https://drive.google.com/a/mail.csuchico.edu/folderview?id=0B83Z8_sNw3KPcnVrYzVFRHUtcHM&usp=sharing)

# Assigned Reading
Afifi: Chapter 9.3, Harrel Ch 2


# Factor variable coding

* Better used term: Indicator variable
* Math notation: **I(gender == "Female")**. 
* A.k.a reference coding
* For a nominal X with K categories, define K indicator variables.
    - Choose a reference (referent) category:
    - Leave it out
    - Use remaining K-1 in the regression.
    - Often, the largest category is chosen as the reference category.


## Example: Binary indicator for gender
Consider the linear model of FEV on gender($x_{1}$), height($x_{2}$) and age($x_{3}$)
where gender interacts with both age and height. In other words, gender changes
the relationship between height and FEV1, and the relationship between age and FEV1.

$$ FEV1 \sim \beta_{0} + \beta_{1}*gender + \beta_{2}*height + \beta_{3}*age + \beta_{4}*gender*height + \beta_{5}*gender*age $$

If we let gender = 0 if the record is on a male, and gender = 1 if the record is
on a female, then the model for males would be:

$$ FEV1 \sim \beta_{0} +  \beta_{2}*height + \beta_{3}*age $$

and the model for females would be:

$$ FEV1 \sim (\beta_{0} + \beta_{1}) + (\beta_{2} + \beta_{4})*height + (\beta_{3}+\beta_{5})*age$$


## Example: Religion against income and depression
Consider a log-linear model for the effect of marital status ($X_2$) on
log income while controlling for age($X_1$). This is called a log-linear
model because the outcome has been log transformed. 

$$ log(Y_i) = \beta_0 + \beta_1*x_1 + \beta_2*x_2 $$

```{r}
dep <- read.table("C:/GitHub/MATH456/data/Depress_020916.txt", sep="\t", header=TRUE)
names(dep) <- tolower(names(dep)) # I hate all captal variable names
levels(dep$marital)
```

Marital status has 5 levels, so we would need 4 indicator variables. 
R always uses the first level of a factor variable as the reference level. 

* Let $x_{2}=1$ when `marital='Married'`, and 0 otherwise,  
* let $x_{3}=1$ when `marital='Never Married'`, and 0 otherwise,  
* let $x_{4}=1$ when `marital='Separated'`, and 0 otherwise,  
* let $x_{5}=1$ when `marital='Widowed'`, and 0 otherwise. 

The mathmatical model would look like: 

$$ log(Y)|X \sim \beta_{0} + \beta_{1}*x_{1} + \beta_{2}x_{2} + \beta_{3}x_{3} + \beta_{4}x_{4} + \beta_{5}x_{5} $$

Two levels of interpretation here. 

1. The outcome is log transformed, so the interpretation has to be 
   back-transformed.   
2. The coefficients for the other levels of the categorical variable
   are in _comparison_ to the reference level. 


**Interpretation of log-linear models** 
Calculate the change in $Y$ that corresponds to a one unit change in $x_1$. 
Since marital status is remaining constant, I will exclude it from the 
calculations below to save space and not to detract from the main point.

Write each equation down

$$ log(Y)|x_1 = \beta_{0} + \beta_{1}x_{1}$$ 
$$ log(Y)|(x_1+1)  = \beta_{0} + \beta_{1}(x_{1}+1)$$ 

Find the difference

$$ (log(Y)|x_1) - (log(Y)|(x_1+1)) = (\beta_{0} + \beta_{1}x_{1}) - (\beta_{0} + \beta_{1}(x_{1}+1))$$ 

and simplify. 

$$ log(\frac{Y|x_1}{Y|x_1+1}) = \beta_{1}$$ 
$$ \frac{Y|x_1}{Y|x_1+1} = e^{\beta_{1}}$$

Each 1-unit increase in $x_{j}$ multiplies the expected value of Y by $e^{\hat{\beta_{j}}}$.  

Interpretation: $100\hat{\beta_{j}}$ is the expected **percentage** change
in $Y$ for a unit increase in $x_{j}$.


The nice thing about factor variables in R, is that the appropriate 
indicator variables are automatically created for you by the linear
model (`lm()`) function.

```{r, echo=1}
summary(lm(log(income) ~ age + marital,data=dep))
cf <- round(exp(coef(summary(lm(log(income) ~ age + marital,data=dep)))),2)
```


* For every year older, a persons income decreases by 1%. (`exp(-0.009)` = `r cf[2,1]`)
* Married individuals have a 52% higher income compared to those who are divorced. (`exp(-0.417)` = `r cf[3,1]`)
* Those who have never been married have  16% lower income compared to those who are divorced. (`exp(-0.183)` = `r cf[4,1]`)
* Separated individuals have 32% lower income compared to those who are divorced. (`exp(-0.394)` = `r cf[5,1]`)
* Widowed individuals have 24% lower income compared to those who are divorced. (`exp(-0.278)` = `r cf[6,1]`)

Other references on how to interpret regression parameters when they have 
been log transformed: 

* http://www.ats.ucla.edu/stat/mult_pkg/faq/general/log_transformed_regression.htm 
* http://www.kenbenoit.net/courses/ME104/logmodels2.pdf

## On Your Own
##### On Your Own

**Create a model to analyze the relationship of education status to depression level as measured by
CESD after controlling for age. Combine all education levels below a HS graduate into one reference 
category called "Up to HS" prior to analysis.**

This is a seemingly simple request, but there are a lot of steps you must do
to correctly analyze this question. 

1. Ensure that you are using the analyzable version of the depression data set. 
   It may be helpful to confirm that your recodes are correct by comparing
   your data mangement code file to mine [dm_depress](../data/dm_depress.html) 
   located on our course website. 
2. Reference your Ch3 homework (or the [solutions](./solutions/ch3_solutions.html))
   if you need help collapsing educational categories. 
3. Ensure that R is treating "Up to HS" as the reference category for eduation level. 
   If it is not, use the `levels` argument of the `factor()` function to reorder
   your factor levels. This is also presented in the Ch3 solutions. 
4. Consider a transformation of `CESD`. Explain and justify using graphical
   measures why you chose to, or chose not to, transform CESD prior to modeling. 
5. Check the model fit by examining the residuals to see if the assumption that 
   $\epsilon_{i} \sim \mathcal{N}(0, \sigma^{2})$ is upheld. 
6. Identify any potential outliers. Explain why you think they are outliers. 
   Examine their standardized residuals and leverage values.If any seem to stand out or 
   have high values for either measure, exclude them from the analysis and re-run the model. 
7. Once you have finalized your model, interpret ALL coefficients in context of the problem. 
   State if any are significantly predictive of the outcome, provide p-values in your conclusion. 
8. Does this model do well at all in predicting CESD? Answer this question using both the
   coefficient of determiniation and the ANOVA test of overall global fit (testing that
   all $\beta$'s are 0)
   


# Splines

## References

* Afifi Section 9.4
* Harrell 2.4.3, pg 39 http://biostat.mc.vanderbilt.edu/tmp/course.pdf 
* https://www.youtube.com/watch?v=o_d4hmKhmsQ
* http://www.r-bloggers.com/thats-smooth/

  



