---
title: "Lec 06: Dimension Reduction using Principal Components"
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
  pdf_document: default
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
library(xtable)
options(xtable.comment = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 9 Overview]](../wk09.html) [[HW Info]](../HW_Info.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)

# Introduction

* Restructure interrelated variables
* Simplify description
* Reduce dimensionality
* Avoid multicollinearity problems in regression

It’s **"Principal"** Components (adjective), not **"Principle"** Components (noun)

From [Grammerist](http://grammarist.com/spelling/principle-principal/): 

> As a noun, principal refers to 
  (1) one who holds a presiding position or rank, and 
  (2) capital or property before interest, 
  and it’s also an adjective meaning 
  (3) first or most important in rank 
> 
> Principle is only a noun. In its primary sense, it refers to a basic truth, law, assumption, or rule.

This third definition (3) is the context in which we will be using this term. 


# Assigned Reading and additional references
* Afifi Chapter 14


# On Your Own
##### On Your Own


    
    
    
    
    