---
title: 'Lec 07: Categorical Data Analysis'
author: "MATH 456 - Spring 2016"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
library(xtable); library(dplyr)
options(xtable.comment = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center') 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 10 Overview]](../wk10.html) [[HW Info]](../HW_Info.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)


# Introduction 
Up until now we have been analyzing continuous outcomes. We will now turn our
focus to methods to analyze categorical data. This will allow us to answer 
questions like the following:

* What proportion of the American public approves of the job the Supreme Court
  is doing?
* The Pew Research Center conducted a poll about support for the 2010 health 
  care law, and they used two forms of the survey question. Each respondent
  was randomly given one of the two questions. What is the difference in the
  support for respondents under the two question orderings?

The methods you learned in previous classes will be useful in these settings. 
For example, sample proportions are well characterized by a nearly normal 
distribution when certain conditions are satisfied, making it possible 
to employ the usual CI and HT tools. In other instances, such as those with 
contingency tables or when sample size conditions are not met, we will use a 
different distribution, though the core ideas remain the same.

Be sure to reference the categorical data sections of the [Data Visualization Tutorial](http://norcalbiostat.github.io/R-Bootcamp/labs/Data_Visualization_Tutorial_Full.html) or 
the [Cookbook for R](http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/) for
more information and guidance on how to plot categorical data correctly. 

#### Assigned Reading
* OpenIntro Lab on categorical data [[HTML]](notes/OI_Inf_for_categorical_data/html)
* OpenIntro Statistics [Free PDF Textbook](https://www.openintro.org/stat/textbook.php) Chapter 6 (6.1-6.4)
* Afifi Chapter 12 (Logistic Regression)

#### Additional References
* Mike Marin: Relative Risk and Odds Ratio https://www.youtube.com/watch?v=V_YNPQoAyCc
* An **excellent** additional textbook reference: _Categorical Data Analysis_ by Alan Agresti
* Using R for Biomedical Statistics http://a-little-book-of-r-for-biomedical-statistics.readthedocs.org/en/latest/src/biomedicalstats.html#
* Graphics
    * **New find!** Graphical Data Analysis with R http://www.gradaanwr.net/content/04-displaying-categorial-data/
    * My Data viz tutorial http://norcalbiostat.github.io/R-Bootcamp/labs/Data_Visualization_Tutorial_Full.html
    * Cookbook for R http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/


#### Spam Data
This set of lecture notes uses the `spam` data set.
```{r}
email <- read.delim("C:/GitHub/MATH456/data/email.txt", header=TRUE, sep="\t")
email <- email %>% mutate(hasnum = ifelse(number %in% c("big", "small"), 1, 0))
```

Two categorical variables of current interest are

* `spam` (0/1 binary indicator if a an email is flagged as spam). Converted into a Ham/Spam factor variable. 
* `number` categorical variable describing the size of the numbers contained in the email.  
    - `none`: No numbers 
    - `small`: Only values under 1 million
    - `big`: A value of 1 million or more
* `hasnum`: 0/1 binary indicator for if the email contains any sized number


#### Chi-Squared Distribution
Much of categorical data analysis uses the $\chi^{2}$ distribution. 

```{r, echo=FALSE, fig.align='center', fig.height=5, fig.width=5}
x <- seq(0,10, by=.01)
df <- c(1,2,3,4,6,8)
plot(c(0,10), c(0,1), type="n", xlab="x", ylab="P(X=x)", axes=FALSE, 
     main="Chi-Squared PDF")
for(k in df){
  lines(x, dchisq(x, k), col=k)
}
legend("topright", col=df, lty=1, legend=df, ncol=2, title="df")
axis(1); axis(2, las=2); box()
```

* The shape is controlled by a degrees of freedom parameter (df)
* Is used in many statistical tests for categorical data. 
* Is always positive (it's squared!)
    - High numbers result in low p-values
* Mathematically connected to many other distributions
    * Special case of the gamma distribution (One of the most commonly used 
      statistical distributions)
    * The sample variance has a $\chi^{2}_{n-1}$ distribution.  
    * The sum of $k$ independent standard normal distributions has a $\chi^{2}_{k}$ distribution. 
    * The ANOVA F-statistic is the ratio of two $\chi^{2}$ distributions divided by 
      their respective degrees of freedom. 


[[top]](lec07_CategoricalDataAnalysis.html)


# Review: Single categorical variable

#### Tables

A table for a single variable is called a _frequency table_. The values 
displayed represent the number of records in the data set that fall into that
particular category.
```{r}
table(email$number)
```

If we replaced the counts with percentages or proportions, the table would be
called a _relative frequency table_.

```{r}
prop.table(table(email$number))
```

We make this output more human readable as percentages by rounding the results
and multiplying by 100.
```{r}
round(prop.table(table(email$number))*100,2)
```

Out of the 3921 emails in this data set, 545 (13.9%) contain big numbers, 
2827 (72.1%) contain small numbers, and 549 (14.0%) contain no numbers. 

#### Barcharts

The most common method to display frequencies in a graphical manner is
with a _bar chart_ (aka barplot or bar graph). One bar per distinct
category with the height of the bar representing the frequency (or percent)
of the data that fall into that category. 

```{r, fig.align='center', fig.width=4, fig.height=4}
ggplot(email, aes(number)) + geom_bar() + ylab("Frequency") + xlab("Size of number in email") + 
  ggtitle("Frequency of emails by size of the number") 
```

#### Inference
This section is considered review and will not be covered directly in 
class. If you have never analyzed proportions in R or it has been a while 
since you took MATH 315 it is advised that you do the following:

* Read Open Intro: Chapter 6.1
* Complete OpenIntro Lab on categorical data [[HTML]](notes/OI_Inf_for_categorical_data/html)

You must have the `openintro` package installed & loaded to have access
to the custom `inference()` function used in the lab.


### Example: Do more than half of the spam emails contain numbers?
We are only interested in the emails that are flagged spam so we filter
the data first. 
```{r}
spam <- email %>% filter(spam==1)
```

#### Using case-level data
Since we created `hasnum` as a binary indicator, we can calculate the
sample proportion as the sample mean. 
```{r}
mean(spam$hasnum)
```
* Construct a 95% CI for the proportion of spam emails with numbers. 
Since this proportion follows the normal model we can conduct a t.test. 
```{r}
t.test(spam$hasnum, alternative = "two.sided", conf.level=.95)
```

A significant majority (59.4%, 95%CI: 54.3-64.4) of emails that are flagged spam
contain numbers (p<.0001).

#### Using summary numbers only
If we only had summary statistics, for example from a published table, 
```{r}
addmargins(table(spam$hasnum))
```
we can still conduct a test to determine if this proportion is 
significantly more than half by calling `prop.test()`. 

```{r}
prop.test(x=218, n=367, p=.5)
```

Significantly more than half (59%, 95%CI 54.2%-64.4%) of emails that were
flagged as spam contain numbers (p<.0004).

[[top]](lec07_CategoricalDataAnalysis.html)


# Difference of two proportions
* OpenIntro Section 6.2

Now let's consider comparisons of proportions in two independent samples. 

**Ex**: Comparison of proportions of head injuries sustained in auto 
accidents by passengers wearing seat belts to those not wearing seat belts. 
You may have already guessed the form of the estimate: $\hat{p}_1 - \hat{p}_2$.

We are not going to go in depth into the calculations for the test statistic 
for a test of the difference in proportions. The OpenIntro textbook explains 
the assumptions and equations very well. Instead we are going to see how to 
use R to perform these calculations for us. 

Since the sample proportion can be calculated as the mean of a binary 
indicator variable, we can use the same `t.test` function in R
to conduct a hypothesis test and create a confidence interval.

### Example 1: Do numbers in emails affect rate of spam? (Case level data) 
If we look at the rate of spam for emails with and without numbers, 
we see that 6% of emails with numbers are flagged as spam compared to 
27% of emails without numbers are flagged as spam. 

```{r}
email %>% group_by(hasnum) %>% summarize(p.spam=round(mean(spam),2))
```

This is such a large difference that we don't really _need_ a statistical
test to tell us that this difference is significant. 
But we will do so anyhow for examples sake. 

1. **State the research question:** Are emails that contain numbers more likely to be spam? 
2. **Define your parameters:**   
Let $p_{nonum}$ be the proportion of emails _without_ numbers that are flagged as spam.  
Let $p_{hasnum}$ be the proportion of emails _with_ numbers that are flagged as spam.  

3. **Set up your statistical hypothesis:**  
$H_{0}: p_{nonum} = p_{hasnum}$   
$H_{A}: p_{nonum} \neq p_{hasnum}$ 

4. **Check assumptions:** Use the pooled proportion $\hat{p}$ to check the success-failure
condition. 
```{r}
p.hat <- mean(email$spam)
p.hat
```

* $\hat{p}*n_{nonum}$ = `p.hat * sum(email$hasnum==0)` = `r p.hat * sum(email$hasnum==0)`
* $\hat{p}*n_{hasnum}$ = `p.hat * sum(email$hasnum==1)` = `r p.hat * sum(email$hasnum==1)`
* $(1-\hat{p})*n_{nonum}$ = `(1-p.hat)* sum(email$hasnum==0)` = `r (1-p.hat)* sum(email$hasnum==0)`
* $(1-\hat{p})*n_{hasnum}$ = `(1-p.hat)* sum(email$hasnum==1)` = `r (1-p.hat) * sum(email$hasnum==1)`

The success-failure condition is satisfied since all values are at least 10, and we can
safely apply the normal model.

5. **Test the hypothesis** by calculating a test statistic and corresponding p-value. 
   Interpret the results in context of the problem. 
```{r}
t.test(spam~hasnum, data=email)
```

Significantly more emails with numbers were flagged as spam compared to emails
without numbers (27.1% versus 6.4% , p<.0001). 

### Example 2: Are mammograms helpful? (Summary numbers only)
**Read Open Intro: 6.2.3**

Test whether there was a difference in breast cancer deaths in the mammogram 
and control groups. By entering in $x$ and $n$ as vectors we can test 
equivalence of these two proportions. The assumptions for using the normal model 
for this test have been discussed in detail in the textbook. 

```{r}
prop.test(x=c(500, 505), n=c(44925, 44910))
```

The interval for the difference in proportions covers zero and the p-value
for the test is 0.894, therefore the proportion of deaths due to breast 
cancer are equal in both groups. There is no indication from this data
that mammograms in addition to regular breast cancer screening, change
the risk of death compared to just the regular screening exams alone. 



[[top]](lec07_CategoricalDataAnalysis.html)

# Contingency tables

* Both the explanatory and the response variables are categorical (Nominal or Ordinal)
* Tables representing all combinations of levels of explanatory and response variables
* A.k.a Two-way tables or _cross-tabs_
* Numbers in table represent Counts of the number of cases in each cell

```{r}
tab <- table(email$spam, email$number)
tab
```

* Row and column totals are called Marginal counts

```{r}
addmargins(tab)
```

## Cell percents
A simple `prop.table()` shows the cell percents. All cells add up to 1. 
```{r}
round(addmargins(prop.table(tab)),3)
```
The `mosaicplot` is a visual representation of these percentages. 
The larger the box the larger the cell proportion. 
```{r}
mosaicplot(tab)
```

We will come back to the usefulness of mosaicplots in analyzing 
contingency tables later. 

## Row percents
Here, the percents add up to 1 across the rows. 
The reference group (denominator) is the row margin.  
This is used when you want to compare the distribution of
percents within each column

**Ex:** How does the distribution of number sizes differ for spam vs non-spam emails? 
```{r}
row.pct <- round(prop.table(tab, 1)*100, 2)
addmargins(row.pct, 2)
```
* 13.9% _of non-spam emails_ contain big numbers
* 40.6% _of spam emails_ contain no numbers
* 74.8% _of non-spam emails_ contain small numbers. 

#### Base graphics
Does "ok", but it depends on how you set up your table (note I had to 
transpose the table `t()` to plot the distribution of numbers within spam).
Also you have to do more work to get a reasonable legend and axes. 
```{r}
barplot(t(tab), beside=TRUE, legend=rownames(t(tab)))
```

#### GGplot2
Requires the data to be aggregated first, then specify the heights
of the bars as a new variable. 
```{r}
library(reshape2)
melted_tab <- melt(tab)
colnames(melted_tab) <- c("Spam", "Number", "count")
melted_tab
ggplot(melted_tab, aes(x = factor(Spam), y= count, fill = Number)) + 
  geom_bar(stat="identity", width=.5, position = "dodge")  + 
  scale_x_discrete("Type of Email", labels=c("Ham", "Spam")) + 
  ylab("Frequency") 
```


## Column percents
Here, the percents add up to 1 down the columns. 
The reference group (denominator) is the column margin.  
This is used when you want to compare the distribution of
the rows within each column. 

**Ex:** How does the distribution of spam differ across 
number sizes?
```{r}
col.pct <- round(prop.table(tab, 2)*100, 2)
addmargins(col.pct,1)
```
* 90.8% _of emails with big numbers_ are not spam
* 27.1% _of emails with no numbers_ are spam
* 94.1% _of emails with small numbers_ are not spam

```{r}
barplot(tab, beside=TRUE, legend=rownames(tab))
```

Once the data is melted, all you have to do is swap x and color 
variables to plot type of email within number size category. 
```{r}
ggplot(melted_tab, aes(x = Number, y= count, fill = factor(Spam))) + 
  geom_bar(stat="identity", width=.5, position = "dodge")  + 
  xlab("Size of number in email") + ylab("Frequency") + 
  scale_fill_discrete("Type of Email", labels=c("Ham", "Spam")) 
```


## Summary / Take home message
It is very important to be clear as to what comparison you (or the researcher)
are interested in making. Sometimes both directions are equally important, 
but often there is one primary direction that is of interest. 


[[top]](lec07_CategoricalDataAnalysis.html)


# Measures of Association
We will consider two measures of association in this class. 

* Relative Risk
* Odds Ratio

These both are calculated on a 2x2 contingency table similar to the following: 

```{r, results='asis'}
nnnn <- matrix(c("$n_{11}$", "$n_{12}$", "$n_{1.}$", 
                 "$n_{21}$", "$n_{22}$", "$n_{2.}$", 
                 "$n_{.1}$", "$n_{.2}$", "$n_{..}$"), nrow=3, byrow=TRUE,  
          dimnames = list(c("Exposed", "Not-Exposed", "Total"), c("Diseased", "Not-Diseased", "Total")))
print(xtable(nnnn, align='cccc'), type='html')
```


Sometimes the cell contents are abbreviated as:
```{r, results='asis'}
abcd <- matrix(c("a", "b", "c", "d"), nrow=2,  
          dimnames = list(c("Exposed", "Not-Exposed"), c("Diseased", "Not-Diseased")))
print(xtable(abcd, align='ccc'), type='html')
```

<br></br>

##### Watch this 6 minute Marin Stats Lecture: https://www.youtube.com/watch?v=V_YNPQoAyCc

## Relative Risk

The **Relative Risk (RR)** or **Risk Ratio** is the ratio of the probability of an 
event occurring in an exposed group compared to the probability of 
an event occurring in a non-exposed group. 

* Asymptotically approaches the OR for small probabilities. 
* Often used in cohort studies and randomized control trials. 

Consider sample proportions Diseases within Exposed and Non-exposed groups.
$$\hat{\pi}_{1} = \frac{n_{11}}{n_{1.}} \qquad \mbox{and} \qquad \hat{\pi}_{2} = \frac{n_{21}}{n_{2.}}$$

The Relative Risk is calculated as 

$$RR = \frac{\hat{\pi}_{1}}{\hat{\pi}_{2}} \qquad \mbox{or} \qquad \frac{a/(a+b)}{c/(c+d)}$$

with variance 
$$V = \frac{1-\hat{\pi}_{1}}{n_{11}} + \frac{1-\hat{\pi}_{2}}{n_{21}}$$


## Odds Ratio 

The **Odds Ratio (OR)** is a way to quantify how strongly the presence
or absence of a characteristic affects the presence or absence of a
second characteristic. 

* Often used in case-control studies
* The main interpretable estimate generated from logistic regression

The **Odds of an event** is the probability it occurs divided by the 
probability it does not occur. 

$$ odds_{1} = \frac{n_{11}/n_{1.}}{n_{12}/n_{1.}} = \frac{n_{11}}{n_{12}} $$
$$ odds_{2} = \frac{n_{21}/n_{2.}}{n_{22}/n_{2.}} = \frac{n_{21}}{n_{22}} $$

The **Odds Ratio** for group 1 compared to group 2 is the ratio of the two 
odds written above: 

$$OR = \frac{odds_{1}}{odds_{2}} = \frac{n_{11}n_{22}}{n_{12}n_{21}} \qquad \mbox{or} \qquad \frac{ad}{bc}$$

with variance $V = n^{-1}_{11} + n^{-1}_{12} + n^{-1}_{21} + n^{-1}_{22}$. 

## Confidence Intervals

Neither the Risk Ratio nor the Odds Ratio are linear functions, so a 95% CI 
for the population estimates are not your typical $\hat{\theta} \pm 1.96\sqrt{Var(\hat{\theta})}$.

Instead they are calculated as the point estimate $\hat{\theta}$ times $e$ 
raised to the $\pm 1.96$ times the standard deviation of the estimate. 

$$(\hat{\theta}e^{-1.96*\sqrt{V}}, \hat{\theta}e^{1.96*\sqrt{V}})$$


### Example: Are emails with numbers in them more likely to be flagged as spam? 
Reconsider the 2x2 table that compares emails flagged as spam to
those containing numbers. 

```{r}
table(email$hasnum, email$spam, dnn=c("Has Number", "Spam"))
```

Note that both the columns and rows are swapped when compared to the a/b/c/d format. 
For ease of interpretation I will recreate the table manually. 
```{r}
tab_sn <- matrix(c(149, 218, 400, 3154), nrow=2, byrow=T, 
                 dimnames = list(c("Has Num", "No Num"), c("Spam", "Ham")))
tab_sn
```

Now I use the `epi.2by2` function contained in the `epiR` package to calculate
the Odds Ratio, the Risk Ratio, and their respective confidence intervals. 
```{r}
library(epiR)
epi.2by2(tab_sn)
```

* Emails containing numbers are 3.6 (3.09, 4.21) times as likely as emails 
  without numbers to be flagged as spam.
* Emails containing numbers have 5.4 (4.27, 6.80) times the odds of being
  flagged as spam compared to emails without numbers in them. 

Both intervals are greater than 1, therefore the event (spam) is statistically
more likely to occur in the exposed group (has num) than in the control (no num) (p<.0001). 
The p-value for the Wald $\chi^{2}$ test is <.0001. 

* Mathematical reference for the Wald test Statistic http://www.statlect.com/Wald_test.htm 


[[top]](lec07_CategoricalDataAnalysis.html)

# Tests of Association

There are three main tests of association for $rxc$ contingency table. 

* Test of Goodness of Fit
* Tests of Independence
* Test of Homogeneity

**Notation**

* $r$ is the number of rows and indexed by $i$
* $c$ is the number of columns and indexed by $j$. 

<br></br>


## Goodness of Fit

* OpenIntro Statistics: Chapter 6.3
* Tests whether a set of multinomial counts is distributed according to a 
  theoretical set of population proportions.
* Does a set of categorical data come from a claimed distribution? 
* Are the observed frequencies consistent with theory? 

$H_{0}$: The data come from the claimed discrete distribution  
$H_{A}$: The data to not come from the claimed discrete distribution. 

<br></br>

## Test of Independence

* OpenIntro Statistics: Chapter 6.4
* Determine whether two categorical variables are associated 
  with one another in the population
      - Ex. Race and smoking, or education level and political affiliation. 
* Data are collected at random from a population and the two 
  categorical variables are observed on each unit. 

$H_{0}: p_{ij} = p_{i.}p_{.j}$   
$H_{A}: p_{ij} \neq p_{i.}p_{.j}$ 

<br></br>

## Test of Homogeneity

* A test of homogeneity tests whether two (or more) sets of multinomial 
  counts come from different sets of population proportions.
* Does two or more sub-groups of a population share the same distribution
  of a single categorical variable?
    - Ex: Do people of different races have the same proportion of smokers?
    - Ex: Do different education levels have different proportions of 
      Democrats, Republicans, and Independents?
* Data on one characteristic is collected from randomly sampling 
  individuals within each subroup of the second characteristic. 

$H_{0}:$ 

$$p_{11} = p_{12} = \ldots = p_{1c}$$
$$p_{21} = p_{22} = \ldots = p_{2c}$$
$$\qquad \vdots $$
$$ p_{r1} = p_{r2} = \ldots = p_{rc}$$


$H_{A}:$ At least one of the above statements is false. 


<br></br>


All three tests use the **Pearsons' Chi-Square** test statistic. 

<br></br>

## Pearsons' Chi-Square
The chi-squared test statistic is the sum of the squared differences 
between the observed and expected values, divided by the expected value. 

### One way table
$$\chi^{2} = \sum^{r}_{i=1} \frac{(O_{i}-E_{i})^2}{E_{i}}$$

* $O_{i}$ observed number of type $i$
* $E_{i}$ expected number of type $i$. Equal to $Np_{i}$ under the null hypothesis
* N is the total sample size
* df = r-1

### Two way tables
$$\chi^{2} = \sum^{r}_{i=1}\sum^{c}_{j=1} \frac{(O_{ij}-E_{ij})^2}{E_{ij}}$$

* $O_{ij}$ observed number in cell $ij$
* $E_{ij} = Np_{i.}p_{.j}$ under the null hypothesis
* N is the total sample size
* df = (r-1)(c-1)

<br></br>

##### Watch this 3 minute Marin Stats Lecture: https://www.youtube.com/watch?v=POiHEJqmiC0

<br></br>

### Conducting these tests in R. 
* Test of equal or given proportions using `prop.test()`
```{r}
prop.test(table(email$number, email$spam))
```

* Chi-squared contingency table tests and goodness-of-fit tests using `chisq.test()`.  
This function can take raw data as input
```{r}
chisq.test(email$number, email$spam)
```
or a table object.
```{r}
chisq.test(tab)
```

### prop.test vs chisq.test()
```{r}
pt.out <- prop.test(table(email$number, email$spam))
cs.out <- chisq.test(tab)
```

* Same calculated test statistic and p-value
```{r}
c(pt.out$statistic, pt.out$p.value)
c(cs.out$statistic, cs.out$p.value)
```
* `prop.test` 
    - has a similar output appearance to other hypothesis tests 
    - shows sample proportions of outcome within each group
* `chisq.test`
    - stores the matricies of $O_{ij}$, $E_{ij}$, the residuals and standardized residuals

```{r}
cs.out$expected
```

_Reminder:_ You can always look at the `names()` of model objects or the help file
to see what information is available for post-analysis investigation. 

### Mosaicplots

* The Pearson $\chi^{2}$ test statistic = Sum of squared residuals. 
* A shaded mosaicplot shows the magnitude of the residuals. 
    * Blue (positive residuals) = More frequent than expected
    * Red (negative residuals) = Less frequent than expected. 

```{r}
mosaicplot(tab, shade=TRUE, main="Association of spam status and number size in emails")
```

There are more spam emails with no numbers, fewer Ham emails with no numbers, 
and fewer spam emails with small numbers than would be expected
if these factors were independent. 
  
* More information on mosaicplots - http://www.datavis.ca/online/mosaics/about.html


# Assumptions and Extensions

* Simple random sample
* Adequate expected cell counts
    - At least 5 in all cells of a 2x2, or at least 80% of cells in a larger table.
    - NO cells with 0 cell count
* Observations are independent

If one or more of these assumptions are not satisfied, other methods
may still be useful. 
* McNemar’s Test for paired or correlated data
* Fishers exact test for when cell sizes are small (<5-10)
* Inter-rater reliability: Concordant and Discordant Pairs


[[top]](lec07_CategoricalDataAnalysis.html)

# On Your Own
For all hypothesis tests you must: 

- Clearly state what your null and alternative hypothesis.  
- Discuss your assumptions
- Show your R code to conduct the test and any data management required prior to the test.
- Write the conclusion of the statistical test in context of the problem. 

##### On Your Own
<!-- TEST OF TWO PROPORTIONS --> 
1. Using a two-sample t-test for proportions on the student `survey` data set
   contained in the `MASS` library, test if males smoke more than females. 
   _Hint: check your output carefully and consider if the difference was 
   taken in the desired direction_
<!--
```{r}
library(MASS)
survey$yessmoker <- ifelse(survey$Smoke !="Never", 1, 0)
t.test(yessmoker ~ Sex , data=survey, alternative = "less")
```
--->

<!-- RISK RATIO --> 
2. Anti-tumor necrosis factor $\alpha$ (TNF$\alpha$) drugs are a class of drugs
   that are commonly used to treat inflammatory conditions such as arthritis. 
   However, these drugs tend to be associated with an increased risk of 
   infectious complications. 
   [Bergstrom, et al (2004)](http://onlinelibrary.wiley.com/doi/10.1002/art.20454/abstract) 
   conducted a study to test the hypothesis that patients on these drugs 
   are at increased risk for coccidiomycosis (a fungal pneumonia). Calculate 
   and interpret the RR with 95% confidence interval.  Here are the tabular results. 
```{r, results='asis'}
tnf <- matrix(c(7, 240, 4, 734), nrow=2, byrow=TRUE,  
          dimnames = list(c("TNF", "Other"), c("COC", "No COC")))
print(xtable(tnf, align='ccc', digits=0), type='html')
``` 

<!-- GOODNESS OF FIT--> 
3. Using the provided bag of M&M's, test the hypothesis that the color 
   distribution has not changed since reported in 2008. 
   https://www.exeter.edu/documents/mandm.pdf
    
<!-- TEST OF INDEPENDENCE --> 
4. In July 2008 the US National Institutes of 
   Health announced that it was stopping a clinical study early because of 
   unexpected results. The study population consisted of HIV-infected women 
   in sub-Saharan Africa who had been given single dose Nevaripine (a 
   treatment for HIV) while giving birth, to prevent transmission of HIV to 
   the infant.  The study was a randomized comparison of continued treatment 
   of a woman (after successful childbirth) with Nevaripine vs. Lopinavir, 
   a second drug used to treat HIV. 240 women participated in the study; 
   120 were randomized to each of the two treatments. Twenty-four weeks after 
   starting the study treatment, each woman was tested to determine if the HIV 
   infection was becoming worse (an outcome called **virologic failure**). 
   Twenty-six of the 120 women treated with Nevaripine experienced virologic failure, 
   while 10 of the 120 women treated with the other drug experienced virologic failure.
   _(Lockman 2007)_
    a. Create a two-way table presenting the results of this study. Include both margins. 
    b. State appropriate hypotheses to test for independence of treatment and virologic failure.
    c. Complete the hypothesis test and state an appropriate conclusion. 
      (Reminder: verify any necessary conditions for the test.)
   
<!--
```{r, echo=FALSE, results='asis'}
hiv <- matrix(c(26, 94, 120, 10, 110, 120, 36, 204, 240), nrow=3, byrow=T,  
       dimnames = list(c("Nevaripine", "Lopinavir", "Total"), c("Yes", "No", "Total")))
print(xtable(hiv, align='cccc'), type='html')

```

(b)~$H_0: p_N = p_L$. There is no difference in virologic failure rates between the Nevaripine and Lopinavir groups. $H_A: p_N \ne p_L$. There is some difference in virologic failure rates between the Nevaripine and Lopinavir groups.
(c)~Random assignment was used, so the observations in each group are independent. If the patients in the study are representative of those in the general population (something impossible to check with the given information), then we can also confidently generalize the findings to the population. The success-failure condition, which we would check using the pooled proportion ($\hat{p} = 36/240 = 0.15$), is satisfied. $Z=3.04$ $\to$ p-value $=0.0024$. Since the p-value is low, we reject $H_0$. There is strong evidence of a difference in virologic failure rates between the Nevaripine and Lopinavir groups do not appear to be independent.}
-->


<!-- INTERPRETING TABLES AND IDENTIFYING INDEPENDENCE OR HOMOGENEITY -->
5. A 2010 survey asked 827 randomly sampled registered voters in California 
   ``Do you support? Or do you oppose? Drilling for oil and natural gas off
   the Coast of California? Or do you not know enough to say?" Below is the
   distribution of responses, separated based on whether or not the respondent
   graduated from college. 
```{r, results='asis'}
oil <- matrix(c(154, 132, 180, 126, 104, 131), nrow=3, byrow=T,  
       dimnames = list(c("Support", "Oppose", "Do not know"), c("Grad", "Non-Grad")))
print(xtable(oil, align='ccc', digits=0), type='html')

```
  a. What percent of college graduates and what percent of the non-college
     graduates in this sample do not know enough to have an opinion on
     drilling for oil and natural gas off the Coast of California?
  b. Create an appropriate graphic to compare the distribution of 
     opinions within college graduates and non-graduates. 
  c. Is this a test of homogeneity or independence? Justify your answer. 

<!-- ODDS RATIO --> 
6. Consider only those participants who have an opinion on off shore drilling. 
   Calculate the Odds Ratio for opposing off shore drilling for college grads
   compared to non-grads. Include a 99% confidence interval and interpret the 
   results in context of the problem.
       
				
    
    