---
title: "Lec 04: Variable Selecdtion"
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
  pdf_document: default
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

Navbar: [[Home]](../index.html) [[Schedule]](../Schedule.html) [[Data]](../data/Datasets.html) [[Week 5 Overview]](../wk05.html) [[HW Rubric]](../admin/rubric.html)  [[Google Group]](https://groups.google.com/forum/#!forum/csuc_stat)  [[Google Drive]](https://drive.google.com/a/mail.csuchico.edu/folderview?id=0B83Z8_sNw3KPcnVrYzVFRHUtcHM&usp=sharing)

# Assigned Reading

# Stuff

## Goodness of fit

### Coefficient of determiniation
Was our model improved by the addition of this variable? Let's check the $R^{2}$. 
Recall this is calculated as the amount of variance explained by the model divided
by the total variance. 
```{r}
summary(aov(mv_model))
(16.05317+5.00380)/(16.05317+5.00380+42.04133)
```
This number is displayed as the **Multiple R-squared** value in the linear model 
output. But what about that other value, the **Adjusted R-squared**? Observe
what happens to these two variables when we put a variable into the model
that is not associated with the outcome, mothers weight.

```{r}
summary(lm(FFEV1 ~ FHEIGHT + FAGE + MWEIGHT, data=fev))
```

Adding more predictors to the model will always increase the $R^{2}$!

**Adjusted R^{2}**
$$ 1 - (1-R^{2})\frac{n-1}{n-p-1} $$




##### On Your Own

