---
title: "Indicators and Splines"
author: "MATH 456 Solutions"
date: "February 26, 2016"
output: html_document
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
```

# Indicators
1. Create a model to analyze the relationship of education status to depression level as measured by
CESD after controlling for age. Combine all education levels below a HS graduate into one reference 
category called "Up to HS" prior to analysis.


# Splines
1. Using the `cars` data set built into R, build a model to predict the
   distance a car takes to stop based on how fast it was going.  
  
```{r}
x6 <- (cars$speed-6)
x6[ x6<0 ] <- 0

x12 <- (cars$speed-12)
x12[ x12<0 ] <- 0

x18 <- (cars$speed-18)
x18[ x18<0 ] <- 0

t(cbind(cars$speed, x6, x12, x18)[8:20,])

fit.lin.spline <- lm(cars$dist ~ cars$speed + x6 + x12 + x18)

ggplot(cars, aes(x=speed, y=dist)) +
    geom_point() + 
    geom_line(aes(x=speed, y=predict(fit.lin.spline)))



```


2. Using the family lung function data, build a model to predict FEV1
   to height for the oldest child. 

```{r}

fev <- read.table("C:/Github/MATH456/data/Lung_020716.txt", sep="\t", header=TRUE)
names(fev) <- tolower(names(fev))
qplot(x=ocheight, y=ocfev1, data=fev, geom='point') + geom_smooth()
```

**SLR**
```{r}
slr <- lm(ocfev1 ~ ocheight, data=fev)
ggplot(fev, aes(x=ocheight, y=ocfev1)) + geom_point() + geom_smooth(se=FALSE, method="lm")
```

**Polynomial**
```{r}
fev$ocheight.sq <- fev$ocheight^2
sqr.model <- lm(ocfev1 ~ ocheight + ocheight.sq, data=fev)

fev$ocheight.cub <- fev$ocheight^3 
cub.model <- lm(ocfev1 ~ ocheight + ocheight.sq + ocheight.cub, data=fev)

# Plot both on same graph
ggplot(fev, aes(x=ocheight, y=ocfev1)) + geom_point() + 
  geom_line(aes(x=ocheight, y=predict(cub.model)), col="blue") + 
  geom_line(aes(x=ocheight, y=predict(sqr.model)), col="red")
```

**Linear Splines**
_Note: A knots at 55 and 64 are just used as demonstration purposes. You need to 
justify your knot placement_

```{r}
# create the knot
gt55 <- (fev$ocheight-55)
gt55[ gt55<0 ] <- 0
gt64 <- (fev$ocheight-64)
gt64[ gt64<0 ] <- 0

# Run the model
lin_spl  <- lm(ocfev1 ~ ocheight + gt55 + gt64, data=fev)

# calculate predictions for the points at knots for plotting
new.dta <- data.frame(ocheight=c(55, 64), gt55=c(0,9), gt64=c(0,0))
pred.pts <- predict(lin_spl, new.dta)

# plot the data and model prediction
ggplot(fev, aes(x=ocheight, y=ocfev1)) + geom_point() + 
    geom_point(x=55, y=pred.pts[1], size=3, col="purple") + 
    geom_point(x=64, y=pred.pts[2], size=3, col="purple") + 
    geom_line(aes(x=ocheight, y=predict(lin_spl)), col="purple")
    
```

Hint: Look at `model.matrix(lm(ocfev1 ~ ocheight + gt55 + gt64, data=fev))` to 
see why the x-values for gt55 is 9 when ocheight is 64. 

**Cubic splines**
```{r}
gt55.3 <- gt55^3
gt64.3 <- gt64^3

cub.spline <- lm(ocfev1 ~ ocheight + ocheight.sq + ocheight.cub + gt55.3 + gt64.3, data=fev)

ggplot(fev, aes(x=ocheight, y=ocfev1)) + geom_point() + 
    geom_line(aes(x=ocheight, y=predict(cub.spline)), col="darkorange")
```

**Natural splines**
```{r}
library(splines)
fit.ns <- lm(ocfev1 ~ ns(ocheight, 3), data=fev )
ggplot(fev, aes(x=ocheight, y=ocfev1)) + geom_point() + 
    geom_line(aes(x=ocheight, y=predict(fit.ns)), col="darkgreen")
```




