---
title: 'Lec 11: Missing Data - Solutions'
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
  pdf_document: default
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(rmarkdown); library(knitr)
options(xtable.comment = FALSE)
opts_chunk$set(warning=FALSE, message=FALSE, fig.height=4, fig.width=4, fig.align='center') 
```

```{r}
lib <- c("VIM", "xtable", "dplyr", "mice", "missForest", "ggplot2","scales", "lattice")
invisible(lapply(lib, library, character.only=T))
```
1. For each distribution listed below, draw a random sample of $n=100$
   and delete 20% of the data completely at random and calculate the bias. 
   State if the observed mean over- or under-estimates the true mean.
    a. $X \sim \mathcal{N}(10,2)$
    b. $Y \sim \mathcal{Poisson}(10)$
    c. $B \sim \mathcal{Binomial}(10, .5)$
    d. $F \sim \mathcal{Binomial}(10, .9)$  
   Does the effect of MCAR on the bias differ for the different distributions? 
2. Repeat #1 but set the missing data mechanism to b NMAR, where $p$ is 
   negatively correlated with the data. 
3. **Using the Parental HIV data set, consider only the following variables:
    `Age`, `Gender`, `livwith`, `BSI_overall`, `Frnds`, and `Hookey`.**
```{r}
hiv <- read.delim("C:/GitHub/MATH456/data/PARHIV_022216.txt")
names(hiv) <- tolower(names(hiv))
hiv <- hiv %>% select(age, gender, livwith, bsi_overall, frnds, hookey) %>%
                  mutate(frnds = as.factor(frnds), livwith=as.factor(livwith), 
                         hookey = hookey-1)
```
**a. What percent of the data set overall is missing?**
```{r}
table(is.na(hiv))
round(mean(is.na(hiv))*100, 2)
```
Only 6 pieces of data (0.4%) are missing. 
    
**b. How much missing data is there per variable?**
```{r}
apply(hiv, 2, function(x) round(sum(is.na(x)),4))
```

**c. Describe the missing data pattern. (_Hint: Use `aggr` from the `VIM` package for part b and c_)**  
```{r, fig.width=6, fig.height=3}
aggr(hiv, col=c("salmon", "slateblue"), numbers=TRUE, sortVars=TRUE, labels=names(hiv), 
     cex.axis=.6, gap=3, ylab=c("% missing data", "Pattern of missing"))
```
  Only `BSI_overall` and `Frnds` have any missing data, with 3 records each, and 
  they are never missing at the same time. 
    
**d. Describe the relationship of missing data between Age and BSI overall.**
```{r}
marginplot(hiv[,c(1,4)])
```
It appears that BSI overall is only missing for youth that are either very young (12), 
  or very old (18). 
    
**4. Multiply impute this data set $m=5$ times.**
```{r}
imp_hiv <- mice(hiv, printFlag=FALSE)
summary(imp_hiv)
```
**a. State the imputation models used for each variable.**  

BSI overall was imputed using predictive mean matching, and `frnds` was
imputed using polytomous logistic regression. These are the defaults for those variable types. 

**5. After controlling for age, who the student lives with, their overall friendships, and
   overall BSI score, what is the effect of gender on the likelihood a student will skip
   school? (I.e. Fit a logistic regression model using `hookey` as the outcome and 
   all other covariates as predictors. Calculate the OR and 95% CI for the effect of gender)**
   
**a. Fit this model on the complete cases (no imputation).**
```{r, results='asis'}
md_model <- glm(hookey ~ age + gender + livwith + bsi_overall + frnds, 
                data=hiv, family="binomial")
print(xtable(summary(md_model), digits=3), type="html")
```
```{r}
exp(c(coef(md_model)[3], confint(md_model)[3,]))
```
Males have 0.70 (.37, 1.32) times the odds of playing hookey compared to females,
a non-significant difference (p=.27).

**b. Fit this model on the multiply imputed data sets and pool the results.**
```{r, results='asis'}
imp_model <- with(imp_hiv, glm(hookey ~ age + gender + livwith + bsi_overall + frnds, 
                               data=hiv, family="binomial"))
print(xtable(summary(pool(imp_model))[,c(1:3, 5:7, 9)], digits=3), type="html")
```
```{r}
exp(summary(pool(imp_model))[3,c(1,6,7)])
```
Males have 0.70 (.37, 1.33) times the odds of playing hookey compared to females, 
a non-significant difference (p=.27). 

**c. Compare these two intervals. Which one is wider? Why?**

The intervals are the same to the second digit. 


**d. Discuss the amount of information lost (in terms of sample size and varance) using the complete case method.**
   
There was no loss of information due to the missing data in this model.
The relative small amount of missing data (n=6) did not affect the estimates or
standard errors. This is a demonstration of how even if the data was NMAR,
the effort required to impute the missing information (even if it wasn't 
much effort at all) does not lead to a noticable change in results. 

Reiteration of two take home messages / common themes throughout the class. 

1. A simpler model is always better
2. When in doubt, do both methods and see if the results actually changed. 

   
         
