---
title: "Multiple Linear Regression (Afifi Ch 7) Solutions to selected problems"
author: "MATH 456 - Spring 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
  pdf_document: default
---
```{r results="asis", echo=FALSE}
source("C:/Github/MATH456/stylesheets/custom.R")
```

```{r, warning=FALSE, message=FALSE}
library(knitr); library(rmarkdown);library(ggplot2);library(xtable)
options(xtable.comment = FALSE) # suppresses the % line from xtable
opts_chunk$set(warning=FALSE, message=FALSE) 

fev <- read.delim("C:/GitHub/MATH456/data/Lung_020716.txt", sep="\t", header=TRUE)
names(fev) <- tolower(names(fev))
ncb <- read.delim("C:/GitHub/MATH456/data/Lung_020716.txt", sep="\t", header=TRUE)
names(fev) <- tolower(names(ncb))
hiv <- read.delim("C:/GitHub/MATH456/data/Lung_020716.txt", sep="\t", header=TRUE)
names(hiv) <- tolower(names(hiv))
```

## 1. Regress FVC on age and height for fathers.  

```{r, results='asis'}
dad_model <- lm(ffvc ~ fage + fheight, data=fev)
print(xtable(summary(dad_model)), type="html")
```


## 2. Confirm the F-test in the summary results is the same as the ANOVA test of global fit. 
Confirm that this F-test in the model results is the correct one to 
   use by manually calculating the F statistic using an ANOVA table. 
   Confirm the degrees of freedom in both the numerator and denominator 
   are correct, as well as the calculation of the p-value.
  
  
## 3. Regress FVC on age and height for mothers.  

```{r, results='asis'}
mom_model <- lm(mfvc ~ mage + mheight, data=fev)
print(xtable(summary(mom_model)), type="html")
```

## 4. Test whether gender changes the effect of age and height on FVC
   
Programmatically extract the coefficients and standard errors
from the model results. 
```{r}
beta_age_male  <- coef(dad_model)[2]; var_beta_age_male <- vcov(dad_model)[2,2]
beta_age_fem   <- coef(mom_model)[2]; var_beta_age_fem  <- vcov(mom_model)[2,2]
beta_ht_male   <- coef(dad_model)[3]; var_beta_ht_fem   <- vcov(mom_model)[3,3]
beta_ht_female <- coef(mom_model)[3]; var_beta_ht_fem   <- vcov(mom_model)[3,3]
```

#### Age
Calculate the test statistic and p-value for $H_{0}: \beta_{1M} = \beta_{1F}$
versus $H_{A}: \beta_{1M} \neq \beta_{1F}$. In other words, is the effect of
age on FEV different for males compared to females. 
```{r}
z = (beta_age_male - beta_age_fem)/(sqrt(var_beta_age_male + var_beta_age_fem))
z
2*(pnorm(z))
```

There appears to be a stronger negative relationship between age and FEV for males
compared to females ($\beta_{1M} = -2.8, \beta_{1F} = -1.8$), but this is not a 
statistically significant difference (p=.31). 

```{r}
confint(dad_model)[2,]
confint(mom_model)[2,]
```

The confidence intervals overlap, so we cannot use this method to determine
if there is a significant difference. 


#### Height




### Alternative
If the data was reshaped into long format, then interaction terms between 
gender and age, and gender and height could be tested individually to answer
the same question. 



## 5. Model birthweight using mothers age, smoking habit and number of visits. 
Interpret the regression coefficients in context of the problem 
and include 95% confidence intervals and p-values in your discussion. 

## 6. Calculate a 95% prediction interval for a 30-year-old smoking mother with 16 visits to the doctor during her pregnancy.
   
## 7. Test for an interaction between smoking habit and the mothers age. 
Include a plot similar to the one shown in the lecture notes to support your findings. 


## 8. Is there a gender difference in when adolescents start smoking adjusting for neighborhood factors? 
Generate a variable that represents the sum
of the variables describing the neighborhood where the adolescent lives
  (`NGHB1-NGHB11`). Is the age at which adolescents start smoking different
   for girls compared to boys, after adjusting for the score describing the 
   neighborhood?**

   