---
title: "M&M Experiment"
author: "Solutions"
date: "April 3, 2016"
output: html_document
---


## Setup / Preparation. 
**1. Is this a Test of Independence, a Test of Homogeneity or a test for Goodness of Fit? Explain.** 

This is a test of goodness of fit because we are comparing an observed distribution to a theoretical or claimed distribution. 

**2. State the null and alternative hypothesis.**

$H_{0}: $  
$H_{A}: $ At least one of these proportions is not as claimed. 

**3. Write the equation to determine the test statistic.**

$$\sum_{i}^{k}\frac{(O_{i}-E_{i})^2}{E_{i}}$$
   
   
**4.	In what case would you reject the null hypothesis? **
You would reject the null hypothesis when your test statistic is large enough. In this case, at the
$\alpha=.05$ significance level and with $k-1=5$ degrees of freedom, "large enough" means observing
a test statistic greater than 11.07. 
```{r}
qchisq(.95, 5)
```

## Calculating probabilities under a $\chi^{2}$ distribution in R. 

**1. Calculate the p-value for your group's bag.**  
```{r}
g1 <- c(12, 10, 12, 4, 8, 10);  g2 <- c(12, 15, 11, 5, 3, 10)
g3 <- c(13, 8, 12, 1, 12, 8);   g4 <- c(13, 7, 13, 8, 4, 11)
p.exp <- c(.24, .2, .16, .14, .13, .13)
```

```{r, echo=FALSE, results='hide', warning=FALSE}
c1 <- chisq.test(g1, p=p.exp);c2 <- chisq.test(g2, p=p.exp)
c3 <- chisq.test(g3, p=p.exp);c4 <- chisq.test(g4, p=p.exp)
```
* Group 1: $\chi^{2}$ = `r round(c1$statistic,2)`.  The p-value is `r round(c1$p.value,2)`
* Group 2: $\chi^{2}$ = `r round(c2$statistic,2)`.  The p-value is `r round(c2$p.value,2)`
* Group 3: $\chi^{2}$ = `r round(c3$statistic,2)`.  The p-value is `r round(c3$p.value,2)`
* Group 4: $\chi^{2}$ = `r round(c4$statistic,2)`.  The p-value is `r round(c4$p.value,2)`


**2. Write a conclusion from this hypothesis test.**

Group 3 should reject the null hypothesis and conclude that the distribution of M & M colors has
changed since 2008. All other groups should not reject, citing insufficient evidence to conclude
that the color distribution has changed. 


## Performing the entire test in R. 

1. Based on the class data what can you conclude about the color distribution of M & M's?

```{r, warning=FALSE}
class <- c(50, 40, 48, 18, 27, 39)
chisq.test(class, p=p.exp)
```

There is sufficient evidence to believe that the Mars company has changed the distribution
of colors in a bag of M&M's since 2008. 

2.	Were your calculated chi square values consistent between your individual and the class data?
Why do you believe this to be true?

Behold the effects of sample size, and/or not accounting for clustering. 


